Supplementary material to `Adaptive group-regularized logistic elastic net regression'

### 2019-04-02 docs/supplement.Rnw
small changes to text of manuscript.Rnw and supplement.Rnw

% vectors and matrices
\newcommand{\bbeta}{\bm{\beta}}
\newcommand{\btau}{\bm{\tau}}
\newcommand{\balpha}{\bm{\alpha}}
\newcommand{\bomega}{\bm{\omega}}
\newcommand{\bOmega}{\bm{\Omega}}
\newcommand{\bSigma}{\bm{\Sigma}}
\newcommand{\bkappa}{\bm{\kappa}}
\newcommand{\btheta}{\bm{\theta}}
\newcommand{\bmu}{\bm{\mu}}
\newcommand{\bpsi}{\bm{\psi}}
\newcommand{\blambda}{\bm{\lambda}}
\newcommand{\bLambda}{\bm{\Lambda}}
\newcommand{\bphi}{\bm{\phi}}
\newcommand{\Y}{\mathbf{Y}}
\newcommand{\y}{\mathbf{y}}
\newcommand{\X}{\mathbf{X}}
\newcommand{\x}{\mathbf{x}}
\newcommand{\I}{\mathbf{I}}
\newcommand{\T}{\mathbf{T}}
\newcommand{\Z}{\mathbf{Z}}
\newcommand{\W}{\mathbf{W}}
\renewcommand{\O}{\mathbf{O}}
\newcommand{\B}{\mathbf{B}}
\renewcommand{\H}{\mathbf{H}}
\newcommand{\U}{\mathbf{U}}
\newcommand{\D}{\mathbf{D}}
\newcommand{\Vm}{\mathbf{V}}
\newcommand{\rv}{\mathbf{r}}
\newcommand{\A}{\mathbf{A}}
\newcommand{\Q}{\mathbf{Q}}
\newcommand{\Sm}{\mathbf{S}}
\newcommand{\0}{\bm{0}}
\newcommand{\tx}{\tilde{\mathbf{x}}}
\newcommand{\hy}{\hat{y}}
\newcommand{\tm}{\tilde{m}}
\newcommand{\tkappa}{\tilde{\kappa}}
\newcommand{\m}{\mathbf{m}}
\newcommand{\C}{\mathbf{C}}
\renewcommand{\v}{\mathbf{v}}
\newcommand{\g}{\mathbf{g}}
\newcommand{\s}{\mathbf{s}}
\renewcommand{\c}{\mathbf{c}}
\newcommand{\ones}{\mathbf{1}}

@@

% settings
\graphicspath{{../figs/}}
\setlength{\evensidemargin}{.5cm} \setlength{\oddsidemargin}{.5cm}
\setlength{\textwidth}{15cm}
\title{Adaptive group-regularized logistic elastic net regression Supplementary 
material}
\date{\today}
\author{Magnus M. M\"unch$^{1,2}$\footnote{Correspondence to: 
\href{mailto:m.munch@vumc.nl}{m.munch@vumc.nl}}, Carel F.W. Peeters$^{1}$, 
Aad W. van der Vaart$^{2}$, and \\ Mark A. van de Wiel$^{1,3}$}

@@

  We start with the computation of (\ref{eq:elboomega}). To this end we define:
  $$
  h(m_i, \omega_i) := \sum_{k=0}^{\infty} (-1)^k \frac{\Gamma(k+m_i}{k + 1}
  \frac{2k +m_i}{\sqrt{2 \pi \omega_i^3}}
  \exp\[-\frac{(2k + m_i)^2}{8 \omega_i} \].
  $$
  Let $\c = \begin{bmatrix} c_1 & \cdots & c_n \end{bmatrix} \tr$. Furthermore, 
  in the following, all expectations and variances are with respect to 
  $Q^{(t)}$. Next we write out the three terms in (\ref{eq:elboomega}):
  \begin{subequations} \label{eq:elboomega2}
    \begin{align}
      \E [\log \L (\y ; \bbeta)] & = \sum_{i=1}^n \log \binom{m_i}{y_i} + 
      \y \tr \X \E [\bbeta] - \m \tr \E \left\{ \log [1 + \exp(\x_i \tr \bbeta)]
      \right\}, \\
      \E [\log \pi(\bomega | \bbeta)] & = \m \tr \E 
      \left\{ \log [1 + \exp(\x_i \tr \bbeta)] \right\} - n \log 2 - 
      \frac{1}{2} \m \tr \X \E [\bbeta] \\
      & \,\,\,\,\, - \sum_{i=1}^n \log \Gamma (m_i) - \frac{1}{2} \sum_{i=1}^n 
      [ ( \x_i \tr \E [\bbeta] )^2 + \x_i \tr \V [\bbeta] \x_i ] 
      \E [\omega_i] \\
      & \,\,\,\,\,+ \sum_{i=1}^n \E 
      \left\{ \log [ h(m_i, \omega_i) ] \right\}, \\
      \E [\log q(\bomega)] & =  \m \tr \log [1 + \exp(\c)] - n \log 2 - 
      \frac{1}{2} \m \tr \c - \sum_{i=1}^n \log \Gamma (m_i) - \frac{1}{2} 
      \sum_{i=1}^n c_i^2 \E [\omega_i] \\
      & \,\,\,\,\, + \sum_{i=1}^n \E \left\{ \log [ h(m_i, \omega_i) ] \right\},
	  \end{align}
  \end{subequations}
  where all mathematical operations on vectors and matrices are element wise. 
  After combining the terms in (\ref{eq:elboomega2}) and substituting the 
  expectations and variances, several terms cancel to
  give:
  \begin{subequations} \label{eq:elboomega3}
	  \begin{align}
	    \E_{Q^{(t)}} & [\log \L (\y ; \bbeta)] + \E_{Q^{(t)}} 
	    [\log \pi(\bomega | \bbeta)] - \E_{Q^{(t)}} [\log q(\bomega)] \\ 
	    & = \sum_{i=1}^n \log \binom{m_i}{y_i} + ( \y - \frac{1}{2} \m) 
	    \tr \X \bmu^{(t)} + \m \tr \left\{ \frac{1}{2} \c^{(t)} - 
	    \log [1 + \exp(\c^{(t)})] \right\} \label{eq:elboomega4}\\
	    & \,\,\,\,\, + \frac{1}{4} \sum_{i=1}^n \frac{m_i}{c_i^{(t)}} 
	    \tanh \(\frac{c_i^{(t)}}{2} \)\left\{ (c_i^{(t)})^2 - 
	    ( \x_i \tr \bmu^{(t)} )^2 - \x_i \tr \bSigma^{(t)} \x_i \tr \right\}.
    \end{align}
  \end{subequations}

@@

  Adding up the terms in (\ref{eq:elboomega3}), (\ref{eq:elbobeta3}), and
  (\ref{eq:elbotau3}) gives the ELBO, at iteration $t$:
  \begin{align*}
    \text{ELBO} (Q^{(t)}) & \propto (\y - \frac{1}{2} \m ) \tr \X \bmu^{(t)} + 
    \m \tr \{ \frac{1}{2} \c^{(t)} - \log [ \exp(\c^{(t)}) + 1 ]\} + \frac{1}{2}
    \sum_{g=1}^G |\G(g)| \log \lambda'^{(t)}_g \\ 
    & + \frac{1}{2}\log |\bSigma^{(t)}| - \frac{\lambda_2}{2} 
    \sum_{g=1}^G \lambda'^{(t)}_g \sum_{j \in \mathcal{G}(g)} 
    \(1 + \frac{\lambda_1}{2 \sqrt{\lambda_2 \chi_j^{(t)} } }\) 
    \((\bmu^{(t)}_j)^2 + \bSigma_{jj}^{(t)}\) \\
    & + \frac{\lambda_1}{4 \sqrt{\lambda_2}} \sum_{j=1}^p \sqrt{\chi_j^{(t)}} -
    \frac{1}{4} \sum_{j=1}^p \log \chi_j^{(t)}.
  \end{align*}
  Here, proportionality is with respect to the variational and penalty 
  parameters.

@@

  Now we rewrite the calculation of $\c = \begin{bmatrix} c_1 & \cdots & c_n
  \end{bmatrix} \tr$ as:
  \begin{align*}
    \c & = \begin{bmatrix} \sqrt{\x_1 \bSigma \x_1 + (\x_1 \tr \bmu)^2} & 
    \cdots & \sqrt{\x_n \bSigma \x_n + (\x_n \tr \bmu)^2} \end{bmatrix} \tr \\
    & = \sqrt{\diag \[\X \bSigma \X \tr\] +  
    \diag \[\X \bmu \bmu \tr \X \tr\]} \\
    & = \sqrt{ [(\X \bSigma) \circ \X ] \cdot \ones_{p \times 1} + 
    (\X \bmu) \circ (\X \bmu)}.
  \end{align*}
  The calculation requires matrix products of complexity $\mathcal{O}(n^2 p)$,
  Hadamard products of $\mathcal{O}(np)$ and matrix inversions of $\mathcal{O}
  (n^3)$, such that the total complexity is of $\mathcal{O}(n^2p)$.

@@

  \section{Simulations}\label{sec:simulations}
  The quality of feature selection is assessed by Cohen's kappa 
  \cite[]{cohen_coefficient_1960}. This kappa is measures similarity between two
  sets of categorical values. One set contains the feature selection indicators:
  $s(j) := \mathbbm{1}\{\hat{\beta}_j \neq 0\}$, while the other set contains
  the true non-zero feature indicators: $t(j) := \mathbbm{1}\{\beta_j \neq 0\}$.
  We calculate Cohen's $\kappa$ as:
	\begin{align*}
	  \kappa & = \frac{f_o - f_e}{1 - f_e}, \text{ where } f_o = p^{-1}
	  \sum_{j=1}^p \mathbbm{1}\{s(j)=t(j)\} \\ & \text{ and } f_e = p^{-2} 
	  \left\{ \(\sum_{j=1}^p s(j)\) \cdot \(\sum_{j=1}^p t(j)\) + 
	  \[\sum_{j=1}^p (1 - s(j)) \] \cdot \[\sum_{j=1}^p (1 - t(j)) \]\right\}.
	\end{align*}
	Here, $f_o$ and $f_e$ denote the frequencies of correctly identified and
	expected correctly identified features, respectively. A positive or negative
	kappa indicate better and worse feature selection than expected by chance,
	respectively. In addition, we evaluate estimation accuracy by mean squared
	error: $\text{MSE} = p^{-1} \sum_{j=1}^p (\beta_j - \hat{\beta}_j)^2$.

  \subsection{Scenario 1}
	In this sparse scenario we assume differential signal between 5 groups of
	features. The average signal in a group increases by a factor 1.6 with group
	number:
	\begin{align*}
	  \bbeta & =  \begin{bmatrix} \smash[b]{\underbrace{ \begin{matrix} 
	  \mathbf{0} & \cdots & \mathbf{0} \end{matrix}}_{150}} 
	  & \smash[b]{\underbrace{ \begin{matrix} \beta_{1}^1 & \cdots & 
	  \beta_{50}^1 \end{matrix}}_{50}} & \cdots & 
	  \smash[b]{\underbrace{ \begin{matrix} \mathbf{0} & \cdots & \mathbf{0}
	  \end{matrix}}_{150}} & \smash[b]{\underbrace{ \begin{matrix} 
	  \beta_{1}^5 & \cdots & \beta^5_{50} \end{matrix}}_{50}} \end{bmatrix} \tr,
	  \\
	  & \, \\
	  \beta_{j}^g & \sim \mathcal{U} (0, b_1 \cdot1.6^{g-1}) \text{, } g=1, \dots,
	  5  \text{, } j=1, \dots, 50,
	\end{align*}
	where $\mathcal{U} (a, b)$ denotes the uniform distribution on $(a,b)$. We set
	$b_1 \approx 0.18$ such that the mean expected signal 
	$\overline{\E(\beta_j^g)}=0.07$. Furthermore, we
	introduce correlation between the features by setting the covariances
	$\bSigma_{ij} = 0.5^{|i-j|}$.The
	model parameters clearly do not follow the relatively heavy-tailed elastic net
	distribution. Because of this mismatch between the assumed and true 
	distributions, we expect poor performance by \texttt{gren}.

@@

	\subsection{Additional results}
	Additional results to MD Section \ref{md-sec:simulations} are given in 
	Figures \ref{fig:lines2_simulations_res1_auc_briers}-
	\ref{fig:lines2_simulations_res5_auc_briers}.

@@

<<heatmap_metabolomics_alzheimer_res1_cor, fig.cap="Correlations between the metabolites.", out.width="50%", fig.asp=1>>=
library(Biobase)
library(lattice)
load("data/ESetMbolCSFPR2.Rdata")
pheno <- pData(ESetMbolCSFPR2)
metabol <- t(exprs(ESetMbolCSFPR2))
metabol.apoe <- metabol[(pheno$D_diag_name=="Probable AD" & 
                           pheno$APOE=="E4YES") |
                          (pheno$D_diag_name=="Subjectieve klachten" & 
                             pheno$APOE=="E4NO"), ]
cor.mat <- cor(scale(metabol.apoe))
levelplot(cor.mat, xlab=NULL, ylab=NULL, scales=list(draw=FALSE))
@

>>>>

% objects
\newcommand{\bbeta}{\bm{\beta}}
\newcommand{\btau}{\bm{\tau}}
\newcommand{\balpha}{\bm{\alpha}}
\newcommand{\bomega}{\bm{\omega}}
\newcommand{\bOmega}{\bm{\Omega}}
\newcommand{\bSigma}{\bm{\Sigma}}
\newcommand{\bkappa}{\bm{\kappa}}
\newcommand{\btheta}{\bm{\theta}}
\newcommand{\bmu}{\bm{\mu}}
\newcommand{\bpsi}{\bm{\psi}}
\newcommand{\blambda}{\bm{\lambda}}
\newcommand{\bLambda}{\bm{\Lambda}}
\newcommand{\bphi}{\bm{\phi}}
\newcommand{\Y}{\mathbf{Y}}
\newcommand{\y}{\mathbf{y}}
\newcommand{\X}{\mathbf{X}}
\newcommand{\x}{\mathbf{x}}
\newcommand{\I}{\mathbf{I}}
\newcommand{\T}{\mathbf{T}}
\newcommand{\Z}{\mathbf{Z}}
\newcommand{\W}{\mathbf{W}}
\renewcommand{\O}{\mathbf{O}}
\newcommand{\B}{\mathbf{B}}
\renewcommand{\H}{\mathbf{H}}
\newcommand{\U}{\mathbf{U}}
\newcommand{\D}{\mathbf{D}}
\newcommand{\Vm}{\mathbf{V}}
\newcommand{\rv}{\mathbf{r}}
\newcommand{\A}{\mathbf{A}}
\newcommand{\Q}{\mathbf{Q}}
\newcommand{\Sm}{\mathbf{S}}
\newcommand{\0}{\bm{0}}
\newcommand{\tx}{\tilde{\mathbf{x}}}
\newcommand{\hy}{\hat{y}}
\newcommand{\tm}{\tilde{m}}
\newcommand{\tkappa}{\tilde{\kappa}}
\newcommand{\m}{\mathbf{m}}
\newcommand{\C}{\mathbf{C}}
\renewcommand{\v}{\mathbf{v}}
\newcommand{\g}{\mathbf{g}}
\newcommand{\s}{\mathbf{s}}
\newcommand{\ones}{\mathbf{1}}

@@

% settings
\graphicspath{{../figs/}}
\setlength{\evensidemargin}{.5cm} \setlength{\oddsidemargin}{.5cm}
\setlength{\textwidth}{15cm}
\title{Supplementary material to 
`Adaptive group-regularized logistic elastic net regression'}
\date{\today}
\author{Magnus M. M\"unch$^{1,2}$\footnote{Correspondence to: 
\href{mailto:m.munch@vumc.nl}{m.munch@vumc.nl}}, Carel F.W. Peeters$^{1}$, 
Aad W. van der Vaart$^{2}$, and \\ Mark A. van de Wiel$^{1,3}$}

@@

  We start with the computation of (\ref{eq:elboomega}). To this end we define:
  $$
  h(m_i, \omega_i) := \sum_{k=0}^{\infty} (-1)^k \frac{\Gamma(k+m_i}{k + 1}
  \frac{2k +m_i}{\sqrt{2 \pi \omega_i^3}}
  \exp\[-\frac{(2k + m_i)^2}{8 \omega_i} \].
  $$
  Let $\mathbf{c} = \begin{bmatrix} c_1 & \cdots & c_n \end{bmatrix} \tr$. Furthermore, 
  in the following, all expectations and variances are with respect to 
  $Q^{(t)}$. Next we write out the three terms in (\ref{eq:elboomega}):
  \begin{subequations} \label{eq:elboomega2}
    \begin{align}
      \E [\log \L (\y ; \bbeta)] & = \sum_{i=1}^n \log \binom{m_i}{y_i} + 
      \y \tr \X \E [\bbeta] - \m \tr \E \left\{ \log [1 + \exp(\x_i \tr \bbeta)]
      \right\}, \\
      \E [\log \pi(\bomega | \bbeta)] & = \m \tr \E 
      \left\{ \log [1 + \exp(\x_i \tr \bbeta)] \right\} - n \log 2 - 
      \frac{1}{2} \m \tr \X \E [\bbeta] \\
      & \,\,\,\,\, - \sum_{i=1}^n \log \Gamma (m_i) - \frac{1}{2} \sum_{i=1}^n 
      [ ( \x_i \tr \E [\bbeta] )^2 + \x_i \tr \V [\bbeta] \x_i ] 
      \E [\omega_i] \\
      & \,\,\,\,\,+ \sum_{i=1}^n \E 
      \left\{ \log [ h(m_i, \omega_i) ] \right\}, \\
      \E [\log q(\bomega)] & =  \m \tr \log [1 + \exp(\mathbf{c})] - n \log 2 - 
      \frac{1}{2} \m \tr \mathbf{c} - \sum_{i=1}^n \log \Gamma (m_i) - \frac{1}{2} 
      \sum_{i=1}^n c_i^2 \E [\omega_i] \\
      & \,\,\,\,\, + \sum_{i=1}^n \E \left\{ \log [ h(m_i, \omega_i) ] \right\},
	  \end{align}
  \end{subequations}
  where all mathematical operations on vectors and matrices are element wise. 
  After combining the terms in (\ref{eq:elboomega2}) and substituting the 
  expectations and variances, several terms cancel to
  give:
  \begin{subequations} \label{eq:elboomega3}
	  \begin{align}
	    \E_{Q^{(t)}} & [\log \L (\y ; \bbeta)] + \E_{Q^{(t)}} 
	    [\log \pi(\bomega | \bbeta)] - \E_{Q^{(t)}} [\log q(\bomega)] \\ 
	    & = \sum_{i=1}^n \log \binom{m_i}{y_i} + ( \y - \frac{1}{2} \m) 
	    \tr \X \bmu^{(t)} + \m \tr \left\{ \frac{1}{2} \mathbf{c}^{(t)} - 
	    \log [1 + \exp(\mathbf{c}^{(t)})] \right\} \label{eq:elboomega4}\\
	    & \,\,\,\,\, + \frac{1}{4} \sum_{i=1}^n \frac{m_i}{c_i^{(t)}} 
	    \tanh \(\frac{c_i^{(t)}}{2} \)\left\{ (c_i^{(t)})^2 - 
	    ( \x_i \tr \bmu^{(t)} )^2 - \x_i \tr \bSigma^{(t)} \x_i \tr \right\}.
    \end{align}
  \end{subequations}

@@

  Adding up the terms in (\ref{eq:elboomega3}), (\ref{eq:elbobeta3}), and
  (\ref{eq:elbotau3}) gives the ELBO, at iteration $t$:
  \begin{align*}
    \text{ELBO} (Q^{(t)}) & \propto (\y - \frac{1}{2} \m ) \tr \X \bmu^{(t)} + 
    \m \tr \{ \frac{1}{2} \mathbf{c}^{(t)} - \log [ \exp(\mathbf{c}^{(t)}) + 
    1 ]\} +
    \frac{1}{2}
    \sum_{g=1}^G |\G(g)| \log \lambda'^{(t)}_g \\ 
    & + \frac{1}{2}\log |\bSigma^{(t)}| - \frac{\lambda_2}{2} 
    \sum_{g=1}^G \lambda'^{(t)}_g \sum_{j \in \mathcal{G}(g)} 
    \(1 + \frac{\lambda_1}{2 \sqrt{\lambda_2 \chi_j^{(t)} } }\) 
    \((\bmu^{(t)}_j)^2 + \bSigma_{jj}^{(t)}\) \\
    & + \frac{\lambda_1}{4 \sqrt{\lambda_2}} \sum_{j=1}^p \sqrt{\chi_j^{(t)}} -
    \frac{1}{4} \sum_{j=1}^p \log \chi_j^{(t)}.
  \end{align*}
  Here, proportionality is with respect to the variational and penalty 
  parameters.

@@

  Now we rewrite the calculation of $\mathbf{c} = \begin{bmatrix} c_1 & 
  \cdots & c_n
  \end{bmatrix} \tr$ as:
  \begin{align*}
    \mathbf{c} & = \begin{bmatrix} \sqrt{\x_1 \bSigma \x_1 + 
    (\x_1 \tr \bmu)^2} & 
    \cdots & \sqrt{\x_n \bSigma \x_n + (\x_n \tr \bmu)^2} \end{bmatrix} \tr \\
    & = \sqrt{\diag \[\X \bSigma \X \tr\] +  
    \diag \[\X \bmu \bmu \tr \X \tr\]} \\
    & = \sqrt{ [(\X \bSigma) \circ \X ] \cdot \ones_{p \times 1} + 
    (\X \bmu) \circ (\X \bmu)}.
  \end{align*}
  The calculation requires matrix products of complexity $\mathcal{O}(n^2 p)$,
  Hadamard products of $\mathcal{O}(np)$ and matrix inversions of $\mathcal{O}
  (n^3)$, such that the total complexity is of $\mathcal{O}(n^2p)$.

@@

  \section{Simulations}\label{sec:simulations}
  \subsection{Scenario 1}
	In this sparse scenario we assume differential signal between 5 groups of
	features. The average signal in a group increases by a factor 1.6 with group
	number:
	\begin{align*}
	  \bbeta & =  \begin{bmatrix} \smash[b]{\underbrace{ \begin{matrix} 
	  \mathbf{0} & \cdots & \mathbf{0} \end{matrix}}_{150}} 
	  & \smash[b]{\underbrace{ \begin{matrix} \beta_{1}^1 & \cdots & 
	  \beta_{50}^1 \end{matrix}}_{50}} & \cdots & 
	  \smash[b]{\underbrace{ \begin{matrix} \mathbf{0} & \cdots & \mathbf{0}
	  \end{matrix}}_{150}} & \smash[b]{\underbrace{ \begin{matrix} 
	  \beta_{1}^5 & \cdots & \beta^5_{50} \end{matrix}}_{50}} \end{bmatrix} \tr,
	  \\
	  & \, \\
	  \beta_{j}^g & \sim \mathcal{U} (0, b_1 \cdot1.6^{g-1}) \text{, } g=1, \dots,
	  5  \text{, } j=1, \dots, 50,
	\end{align*}
	where $\mathcal{U} (a, b)$ denotes the uniform distribution on $(a,b)$. We set
	$b_1 \approx 0.18$ such that the mean expected signal 
	$\overline{\E(\beta_j^g)}=0.07$. Furthermore, we
	introduce correlation between the features by setting the covariances
	$\bSigma_{ij} = 0.5^{|i-j|}$.The
	model parameters clearly do not follow the relatively heavy-tailed elastic net
	distribution. Because of this mismatch between the assumed and true 
	distributions, we expect poor performance by \texttt{gren}.

@@

	\subsection{Additional results}
	Results for different values of the tuning parameter $\alpha$ than presented
	in MD Section \ref{md-sec:simulations} are given in 
	Figures \ref{fig:lines2_simulations_res1_auc_briers}-
	\ref{fig:lines2_simulations_res5_auc_briers}. Overall, these results follow 
	the same pattern as the ones presented in MD Section \ref{md-sec:simulations}.

@@

<<heatmap_metabolomics_alzheimer_res1_cor, fig.cap="Correlations between the metabolites.", out.width="50%", fig.asp=1>>=
library(lattice)
load("data/metabolomics_alzheimer_data.Rdata")
cor.mat <- cor(scale(metabol.apoe))
levelplot(cor.mat, xlab=NULL, ylab=NULL, scales=list(draw=FALSE))
@
